---
title: "Correcting Partial Verification Bias in Diagnostic Accuracy Studies Using R"
author: "Wan Nor Arifin"
date: "Nov 19 2020"
output:
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    theme: readable
    highlight: default
    #code_folding: hide
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Verification Biases

## No Bias

![](allV.svg)

## Partial Verification Bias

![](PVB.svg)

## Differential Verification Bias

![](DVB.svg)

# Correction Methods

## Overview

![](VBcorrection.svg)

## Data

### Data set
Data, 2003a Kosinski & Barnhart:

1. X1 = Gender: 1 = male, 0 = female.
2. X2 = Stress mode: 1 = dipyridamole, 0 = exercise.
3. X3 = Age: 1 = age ≥ 60 years, 0 = age < 60 years.
4. T = SPECT thallium test: 1 = positive, 0 = negative.
5. D = CAD (disease) status: 1 = yes, 0 = no; NA = not verified.

```{r}
cad = read.csv("https://wnarifin.github.io/data/2003kosinski_cad.csv")
str(cad)
```

### Setup
```{r}
# Verification status
cad$R = 1  # R = Verified: 1 = yes, 0 = no
cad[is.na(cad$D), "R"] = 0

# Incomplete data with NA recoding
cad_ = cad
cad_[is.na(cad_$D), "D"] = -1  # recode NA as -1, this will be -1*0=0 for unverified. Else NA*0=NA.
```

### Library
```{r}
library(prediction)
library(mice)
```

## Complete-Case Analysis
$$
\widehat{Sensitivity}_{CC} = P(T=1|D=1)
$$

$$
\widehat{Specificity}_{CC} = P(T=0|D=0)
$$

```{r}
tbl = table(cad[, c("T", "D")]); tbl
sn0 = tbl[2,2]/sum(tbl[,2]); sn0
sp0 = tbl[1,1]/sum(tbl[,1]); sp0
```

## Begg & Greenes

$$
\begin{split}
\widehat{Sensitivity}_{BG} & = P(T=1|D=1) \\ 
& = \frac{P(T=1)P(D=1|T=1,V=1)}{\sum_{T}{P(T)P(D=1|T,V=1)}}
\end{split}
$$

$$
\begin{split}
\widehat{Specificity}_{BG} & = P(T=0|D=0) \\
& = \frac{P(T=0)P(D=0|T=0,V=1)}{\sum_{T}{P(T)P(D=0|T,V=1)}}
\end{split}
$$

### B&G, Count Method in Begg & Greenes 1983
```{r}
tbl_ = table(cad_[, c("T", "D")]); tbl_  # -1 column -- unverified cases
upper = (sum(cad_$T)/nrow(cad_)) * (sum(cad_$D*cad_$T*cad_$R)/sum(cad_$T*cad_$R))
lower = (sum(1-cad_$T)/nrow(cad_)) * (sum(cad_$D*(1-cad_$T)*cad_$R)/sum((1-cad_$T)*cad_$R))
sn1 = upper / (upper+lower); sn1
upper1 = (sum(1-cad_$T)/nrow(cad_)) * (sum((1-cad_$D)*(1-cad_$T)*cad_$R)/sum((1-cad_$T)*cad_$R))
lower1 = (sum(cad_$T)/nrow(cad_)) * (sum((1-cad_$D)*cad_$T*cad_$R)/sum(cad_$T*cad_$R))
sp1 = upper1 / (upper1+lower1); sp1
```

### B&G, Regression Method in Alonzo 2005
```{r}
model1 = glm(D ~ T, data = cad, family = "binomial")  # Modelled using complete data
summary(model1)
preds = prediction(model1, cad_)$fitted  # Predicted on incomplete data
sn2 = sum(cad_$T*preds) / sum(preds); sn2            # P(T=1|D=1)
sp2 = sum((1-cad_$T)*(1-preds)) / sum(1-preds); sp2  # P(T=0|D=0)
```

## Multiple Imputation
MI by Logistic Regression. Also refer to codes by deGroot et al 2008 for Predictive Mean Matching.

```{r results='hide'}
m = 10  # number of imputed data sets
seednum = 12345
cad1 = cad[, c("T","D")]  # select only T & D, else MI use all variables for imputation
cad1$D = as.factor(cad1$D); data_impute = mice(cad1, m = m, method = "logreg", seed = seednum)  # logistic regression
```

```{r}
# results dependent on seed & method, seems pmm less variant than logreg
data_impute$method
cad_mi = complete(data_impute, action = "repeated")  # MI data
names(cad_mi)
```

```{r}
# loop to calculate sn & spe
sn3m = sp3m = rep(NA, m)
for (i in 1:m) {
  tbl = table(T=cad_mi[,1],D=cad_mi[,m+1])
  sn3m[i] = tbl[2,2]/sum(tbl[,2])
  sp3m = tbl[1,1]/sum(tbl[,1])
}
sn3 = mean(sn3m); sn3
sp3 = mean(sp3m); sp3
```

## EM Algorithm
By running multiple cycles of weighted logistic regression models on pseudodata -- Konsinski & Barnhart, 2003. They considered selection model for joint probability of $V$, $T$ and $D$ given covariates $\mathbf{X}$ as follows:
$$
\label{eq:kos1}
P(V,T,D|\mathbf{X}) = P(D|\mathbf{X}) \times P(T|D,\mathbf{X}) \times P(V|T,D,\mathbf{X})
$$

They parameterize the components in the equation with three separate logistic regression models:

Disease component:
$$logit\ P(D_i=1|\mathbf{x}_i)=\mathbf{z}_{0i}^{'}\mathbf{\alpha}$$

Diagnostic test component: 
$$logit\ P(T_i=1|D_i,\mathbf{x}_i)=\mathbf{z}_{1i}^{'}\mathbf{\beta}$$

Missing data mechanism component: 
$$logit\ P(R_i=1|D_i,T_i,\mathbf{x}_i)=\mathbf{z}_{2i}^{'}\mathbf{\gamma}$$
The method is implemented as follows:

1. Pseudodata are created with $n+u$ rows (observations). The first $n-u$ rows are for verified patients ($V=1$), while the remaining $2u$ rows are double-stacked data for unverified patients with missing disease status, $D$. For the first $u$ stack $D$ is set as $D=0$, while for the second $u$ stack $D$ is set as $D=1$.
2. Weighted logistic regression models are fit iteratively (M-step), with weights $w_k(\theta)$ updated in each iteration at the preceding E-step.

The weights $w_k(\theta)$ are defined as follows:
$$
w_{k}(\theta) = \begin{cases}
1 & \text{for}\ k=1,...,n-u\\
p_{k}(\theta)/\{p_k(\theta) + p_{k+U}(\theta) \} & \text{for}\ k=n-u+1,...,n\\
1-w_{k-u}(\theta) & \text{for}\ k=n+1,...,n+u
\end{cases}
$$
where $p_{k}(\theta) = \prod_{m=0}^2 \{p_{mk}(\theta)\}^{y_{mk}} \{1-p_{mk}(\theta)\}^{1-y_{mk}}$. $p_{mk}$ can be easily obtained from the respective logistic regression models for each observation at each iteration using standard logistic regression module by obtaining the probability of the outcomes (i.e. $D=1$, $T=1$ or $V=1$) for each of the patients.

### Data preparation
```{r}
# pseudo-data
cad$R = 1  # R = Verified: 1 = yes, 0 = no
cad[is.na(cad$D), "R"] = 0
cad_pseudo = rbind(cad, cad[cad$R == 0, ])  # replicate U observation
str(cad_pseudo)
sum(is.na(cad_pseudo$D))  # 2217*2 = 4434
table(cad_pseudo$T)
# create 0, 1 for 2U rows
cad_pseudo[(sum(cad$R)+1):nrow(cad), "D"] = 0        # 1st U rows
cad_pseudo[(nrow(cad)+1):nrow(cad_pseudo), "D"] = 1  # 2nd U rows
# view
head(cad_pseudo)
head(cad_pseudo[(sum(cad$R)+1):nrow(cad),])
head(cad_pseudo[(nrow(cad)+1):nrow(cad_pseudo),])
n_u = nrow(cad_pseudo)  # total cases + U
# index k
index_1 = which(cad$R == 1)  # verified
index_2 = (sum(cad$R)+1):nrow(cad)  # unverified U
index_3 = (nrow(cad)+1):nrow(cad_pseudo)  # unverified 2U
```

### Logistic regression models
Components: a = disease, b = diagnostic, c = missing data mechanism

```{r}
# initialize values
weight_k = weight_ka = weight_kb = weight_kc = rep(1, nrow(cad_pseudo))  # init weight = 1 for all
coef_1 = vector("list", 0)
max_t = 500  # may increase
```

```{r warning=FALSE}
# EM Algorithm Iteration
for (t in 1:max_t) {
  # a -- P(D)
  model1a = glm(D ~ 1, data = cad_pseudo, family = "binomial", weight = weight_k)
  su_model1a = summary(model1a)
  coef_1a = su_model1a$coefficients
  fitted_pa = model1a$fitted.values
  # b -- P(T|D)
  model1b = glm(T ~ D, data = cad_pseudo, family = "binomial", weight = weight_k)
  su_model1b = summary(model1b)
  coef_1b = su_model1b$coefficients
  fitted_pb = model1b$fitted.values
  # c -- P(R|T)
  model1c = glm(R ~ T, data = cad_pseudo, family = "binomial", weight = weight_k)
  su_model1c = summary(model1c)
  coef_1c = su_model1c$coefficients
  fitted_pc = model1c$fitted.values
  # all
  coef_1 = append(coef_1, list(list(coef_1a, coef_1b, coef_1c)))
  # weight
  fitted_ps = list(fitted_pa, fitted_pb, fitted_pc)
  ys = list(model1a$y, model1b$y, model1c$y)
  p0 = (fitted_ps[[1]]^ys[[1]]) * ((1 - fitted_ps[[1]])^(1 - ys[[1]]))  # P(D)
  p1 = (fitted_ps[[2]]^ys[[2]]) * ((1 - fitted_ps[[2]])^(1 - ys[[2]]))  # P(T|D)
  p2 = (fitted_ps[[3]]^ys[[3]]) * ((1 - fitted_ps[[3]])^(1 - ys[[3]]))  # P(R|T)
  pk = p0 * p1 * p2  # P(R,T,D|X)
  weight_k[index_2] = pk[index_2] / (pk[index_2] + pk[index_3])
  weight_k[index_3] = 1 - weight_k[index_2]
  # next iteration
  t = t + 1
}
```

### Model summary & estimates
```{r}
coef_1[[max_t]]
sn4 = predict(model1b, list(D=1), type="response"); sn4      # P(T=1|D=1)
sp4 = 1 - predict(model1b, list(D=0), type="response"); sp4  # P(T=0|D=0)
```

## Compare the models
```{r}
snsp = c(sn0,sp0,sn1,sp1,sn2,sp2,sn3,sp3,sn4,sp4)
snsp = matrix(snsp, ncol = 2, nrow = length(snsp)/2, byrow = T)
rownames(snsp) = c("Complete-case","B&G Count","B&G Regression","Multiple Imputation","EM Algorithm")
colnames(snsp) = c("Sensitivity","Specificity")
tbl_compare = round(snsp, 3)
knitr::kable(tbl_compare)
```
